---
title: "Project_dataclean"
author: "Zhan_Zhao"
date: "2026-3-31"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
    code_folding: hide
---

## Task Description

This R Markdown document reads, cleans, and explores the datasets used for the final project.  
It focuses on diagnosed diabetic eye disease in a commercially insured population and examines variation across risk-factor subgroups.  
It also incorporates state-level behavioral indicators from BRFSS to provide additional context.  


The main tasks are:

1) load and inspect the primary and supplementary datasets  
2) identify and define the primary diabetic eye disease outcome  
3) clean and simplify both datasets for state-level analysis  
4) select behavioral indicators from BRFSS and reshape them into a state-year summary file  
5) join the two datasets at the state-year level  
6) create descriptive summaries and exploratory visualizations  
7) document preliminary patterns for later discussion of limitations and future work  


# Initialize
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      fig.width = 8,
                      fig.height = 6) 
#```{r message = TRUE}
# code that you want to run and then the message to display
library(tidyverse)
library(readr)
library(janitor)
```

# define data directories
```{r}
project.dir <- "/Users/zhaozhan/Project"
data.dir <- file.path(project.dir, "01 Data")
output.dir <- file.path(project.dir, "02 Output")

mscan.file <- file.path(
  data.dir,
  "Commercial_Medical_Insurance_(MSCANCC)_-_Vision_and_Eye_Health_Surveillance_20260331.csv"
)

brfss.file <- file.path(
  data.dir,
  "Nutrition,_Physical_Activity,_and_Obesity_-_Behavioral_Risk_Factor_Surveillance_System_20260331.csv"
)
```

# Load and Inspect the Primary Dataset
```{r Load_and_inspect_primary_dataset}
#load_mscan
mscan <- read_csv(mscan.file, show_col_types = FALSE) |>
  clean_names()

glimpse(mscan)
#insert_mscan_questions
mscan |>
  count(topic, sort = TRUE)

mscan |>
  count(question, sort = TRUE) |>
  print(n = 50)

#insert_diabetic_eye_question
mscan |>
  filter(question == "Annual prevalence of diagnosed diabetic eye diseases") |>
  count(response, sort = TRUE)

# Build the Primary Analytic Subset

diab_eye_state <- mscan |>
  filter(
    question == "Annual prevalence of diagnosed diabetic eye diseases",
    response == "All Diabetic eye diseases",
    geographic_level == "State",
    age == "All ages",
    sex == "Both sexes",
    race_ethnicity == "All races",
    risk_factor == "Diabetes",
    risk_factor_response == "Yes"
  ) |>
  transmute( 
    # select variables and rename it
    year_start,
    state_abbr,
    location_desc,
    response,
    diab_eye_prev = data_value,
    diab_eye_lcl = low_confidence_limit,
    diab_eye_ucl = high_confidence_limit,
    sample_size,
    numerator
  )

glimpse(diab_eye_state)

head(diab_eye_state)

diab_eye_state |>
  count(year_start, sort = TRUE)

diab_eye_state |>
  count(state_abbr, sort = TRUE)
```

# Load and Inspect the Supplementary Dataset
```{r Load_and_inspect_supplementary_dataset}
#load brfss
brfss <- read_csv(brfss.file, show_col_types = FALSE) |>
  clean_names()

glimpse(brfss)

#insert brfss
brfss |>
  count(topic, sort = TRUE)

brfss |>
  count(question, sort = TRUE) |>
  print(n = 100)

#Select BRFSS behavioral indicators
brfss_selected <- brfss |>
  filter(
    question %in% c(
      "Percent of adults aged 18 years and older who have obesity",
      "Percent of adults who engage in no leisure-time physical activity"
    )
  )

brfss_selected |>
  count(question, stratification_category1, sort = TRUE) |>
  print(n = 100)

#Bulid brafss_state
brfss_state <- brfss_selected |>
  filter(stratification_category1 == "Total") |>
  select(
    year_start, location_abbr, location_desc,
    question, data_value,
    low_confidence_limit, high_confidence_limit,
    sample_size
  )

brfss_wide <- brfss_state |>
  mutate(
    behavior_indicator = case_when(
      question == "Percent of adults aged 18 years and older who have obesity" ~ "obesity_pct",
      question == "Percent of adults who engage in no leisure-time physical activity" ~ "inactive_pct",
      TRUE ~ NA_character_
    )
  ) |>
  select(year_start, location_abbr, behavior_indicator, data_value) |>
  pivot_wider(
    names_from = behavior_indicator,
    values_from = data_value
  )

glimpse(brfss_wide)
head(brfss_wide)

```

# Join the Two Datasets
```{r Join_the_two_datasets}
# Join data
joined_data <- diab_eye_state |>
  left_join(
    brfss_wide,
    by = c("state_abbr" = "location_abbr", "year_start" = "year_start")
  )

glimpse(joined_data)
head(joined_data)

# Join quality summary
joined_data |>
  summarise(
    n_rows = n(),
    n_states = n_distinct(state_abbr),
    n_years = n_distinct(year_start),
    missing_obesity = sum(is.na(obesity_pct)),
    missing_inactive = sum(is.na(inactive_pct)),
    complete_cases = sum(!is.na(diab_eye_prev) & !is.na(obesity_pct) & !is.na(inactive_pct))
  )

# Check unmatched or incomplete records
joined_data |>
  filter(is.na(obesity_pct) | is.na(inactive_pct)) |>
  select(year_start, state_abbr, location_desc, obesity_pct, inactive_pct)

# Create complete-case analytic dataset
analysis_data <- joined_data |>
  drop_na(diab_eye_prev, obesity_pct, inactive_pct)

glimpse(analysis_data)
head(analysis_data)
```

# Data display
```{r Plot}
# Helper function to compute regression labels for plot annotations

build_lm_label <- function(df, x_var, y_var = "diab_eye_prev") {
  model_data <- df |>
    select(all_of(c(x_var, y_var))) |>
    drop_na()
  fit <- lm(reformulate(x_var, y_var), data = model_data)
  fit_summary <- summary(fit)
  # coef
  fit_coef <- coef(fit_summary)
  # 95% CI
  fit_ci <- confint(fit)
  slope_name <- x_var
  #slope value
  slope_est <- fit_coef[slope_name, "Estimate"]
  # P-value
  slope_p <- fit_coef[slope_name, "Pr(>|t|)"]
  # low ci
  slope_ci_low <- fit_ci[slope_name, 1]
  #high ci
  slope_ci_high <- fit_ci[slope_name, 2]
  r_sq <- fit_summary$r.squared
  paste0(
    "R² = ", sprintf("%.3f", r_sq),
    "\nSlope = ", sprintf("%.3f", slope_est),
    "\n95% CI = [", sprintf("%.3f", slope_ci_low), ", ", sprintf("%.3f", slope_ci_high), "]",
    "\np-value = ", format.pval(slope_p, digits = 3, eps = 0.001)
  )
}

# Build model labels
obesity_label <- build_lm_label(analysis_data, "obesity_pct")
inactive_label <- build_lm_label(analysis_data, "inactive_pct")

# Fit simple linear regression models
obesity_fit <- lm(diab_eye_prev ~ obesity_pct, data = analysis_data)
inactive_fit <- lm(diab_eye_prev ~ inactive_pct, data = analysis_data)

# Figure 1: State-level prevalence
p_state <- ggplot(
  analysis_data,
  aes(x = reorder(state_abbr, diab_eye_prev), y = diab_eye_prev, fill = diab_eye_prev)
) +
  geom_col(color = "gray25", alpha = 0.9) +
  coord_flip() +
  scale_fill_gradient(
    low = "#9ecae1",
    high = "#08519c",
    name = "Prevalence (%)"
  ) +
  labs(
    title = "State-Level Prevalence of Diagnosed Diabetic Eye Disease",
    x = "State",
    y = "Crude prevalence (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

p_state

ggsave(
  file.path(output.dir, "figure1_state_prevalence.png"),
  plot = p_state,
  width = 9,
  height = 7,
  dpi = 300
)

# Figure 2: Obesity and diabetic eye disease
p_obesity <- ggplot(analysis_data, aes(x = obesity_pct, y = diab_eye_prev)) +
  geom_point(color = "#2b8cbe", size = 2.5, alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE, color = "#de2d26", fill = "#fcae91", linewidth = 1) +
  annotate(
    "label",
    x = Inf, y = Inf,
    label = obesity_label,
    hjust = 1.05,
    vjust = 1.1,
    size = 3.3,
    label.size = 0.2
  ) +
  labs(
    title = "State-Level Obesity and Diagnosed Diabetic Eye Disease",
    x = "Obesity prevalence (%)",
    y = "Diagnosed diabetic eye disease prevalence (%)",
    subtitle = "Linear fit with 95% confidence interval"
  ) +
  theme_minimal()

p_obesity

ggsave(
  file.path(output.dir, "figure2_obesity_association.png"),
  plot = p_obesity,
  width = 8,
  height = 6,
  dpi = 300
)

# Short interpretation for obesity model
obesity_coef <- summary(obesity_fit)$coefficients
obesity_ci <- confint(obesity_fit)

cat(
  "**Interpretation (Obesity):** A 1 percentage-point increase in state-level adult obesity prevalence was associated with an estimated ",
  sprintf("%.3f", obesity_coef["obesity_pct", "Estimate"]),
  " percentage-point change in diagnosed diabetic eye disease prevalence (95% CI: ",
  sprintf("%.3f", obesity_ci["obesity_pct", 1]), ", ",
  sprintf("%.3f", obesity_ci["obesity_pct", 2]),
  "; p-value ", format.pval(obesity_coef["obesity_pct", "Pr(>|t|)"], digits = 3, eps = 0.001),
  "). This ecological association should not be interpreted as an individual-level causal relationship.\n\n",
  sep = ""

)

# Figure 3: Physical inactivity and diabetic eye disease
p_inactive <- ggplot(analysis_data, aes(x = inactive_pct, y = diab_eye_prev)) +
  geom_point(color = "#31a354", size = 2.5, alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE, color = "#756bb1", fill = "#dadaeb", linewidth = 1) +
  annotate(
    "label",
    x = Inf,
    y = Inf,
    label = inactive_label,
    hjust = 1.05,
    vjust = 1.1,
    size = 3.3,
    label.size = 0.2
  ) +
  labs(
    title = "Physical Inactivity and Diagnosed Diabetic Eye Disease",
    x = "No leisure-time physical activity (%)",
    y = "Diagnosed diabetic eye disease prevalence (%)",
    subtitle = "Linear fit with 95% confidence interval"
  ) +
  theme_minimal()

p_inactive
ggsave(
  file.path(output.dir, "figure3_inactivity_association.png"),
  plot = p_inactive,
  width = 8,
  height = 6,
  dpi = 300
)

# Short interpretation for physical inactivity model
inactive_coef <- summary(inactive_fit)$coefficients
inactive_ci <- confint(inactive_fit)

cat(
  "**Interpretation (Physical inactivity):** A 1 percentage-point increase in state-level physical inactivity prevalence was associated with an estimated ",
  sprintf("%.3f", inactive_coef["inactive_pct", "Estimate"]),
  " percentage-point change in diagnosed diabetic eye disease prevalence (95% CI: ",
  sprintf("%.3f", inactive_ci["inactive_pct", 1]), ", ",
  sprintf("%.3f", inactive_ci["inactive_pct", 2]),
  "; p-value ", format.pval(inactive_coef["inactive_pct", "Pr(>|t|)"], digits = 3, eps = 0.001),
  "). This ecological association should not be interpreted as an individual-level causal relationship.\n",
  sep = ""
)
```

# Correlation Analysis & Heatmap
```{r Correlation-Heatmap}
# Correlation heatmap for final analytic variables

#Create complete-case data for correlation analysis

cor_data <- analysis_data |>
  select(
    diab_eye_prev,
    obesity_pct,
    inactive_pct
  ) |>
  drop_na()

#Rename Variables for clearer plot labels
cor_data_labeled <- cor_data |>
  rename(
    `Diagnosed diabetic eye disease` = diab_eye_prev,
    `Adult obesity` = obesity_pct,
    `Physical inactivity` = inactive_pct
  )

#Create Pearson correlation matirx

corr_mat <- round(
  cor(
    cor_data_labeled,
    use = "complete.obs",
    method = "pearson"),
  2
)

corr_mat

#convert correlation matrix to long format
corr_heatmap_data <- as.data.frame(as.table(corr_mat)) |>
  rename(
    variable_1 = Var1,
    variable_2 = Var2,
    correlation = Freq
  )

#Figure4: Correlation heatmap
p_corr_heatmap <- ggplot(
  corr_heatmap_data,
  aes(
    x = variable_1,
    y = variable_2,
    fill = correlation)
  )+
    geom_tile(color = "white")+
    geom_text(
      aes(
        label = sprintf("%.2f",correlation)),
        color = "black",
        size = 4
      )+
    scale_fill_gradient2(
      low = "#b2182b",
      mid = "white",
      high = "#2166ac",
      midpoint = 0,
      limits = c(-1,1),
      name = "Pearson r"
    )+
    labs(
      title = "Correlation Heatmap of Diabetic Eye Disease and Behavioral Indicators",
      x = "",
      y = ""
    )+
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid = element_blank()
  )

p_corr_heatmap

ggsave(
  file.path(output.dir, "figure4_correlation_heatmap.png"),
  plot = p_corr_heatmap,
  width = 8,
  height = 6,
  dpi = 300
)

# Correlation tests with p-values

cor_obesity <- cor.test(
  analysis_data$diab_eye_prev,
  analysis_data$obesity_pct,
  method = "pearson"
)

cor_inactive <- cor.test(
  analysis_data$diab_eye_prev,
  analysis_data$inactive_pct,
  method = "pearson"
)

cor_summary <- tibble(
  predictor = c("Adult obesity", "Physical inactivity"),
  correlation_r = c(
    unname(cor_obesity$estimate),
    unname(cor_inactive$estimate)
  ),
  p_value = c(
    cor_obesity$p.value,
    cor_inactive$p.value
  )
)

cor_summary

write_csv(
  cor_summary,
  file.path(output.dir, "correlation_summary.csv")
)
```





